---
title: "ellipsenm: An R package for ecological niche's characterization using ellipsoids"
author: "Marlon E. Cobos, Luis Osorio-Olvera, Jorge Soberón, Laura Jiménez, A. Townsend Peterson, Vijay Barve, and Narayani Barve"
output:
  github_document:
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

<br>

**This repository is for the project "Grinnellian ecological niches and ellipsoids in R" developed during the program GSoC 2019.** 

<br>

## Project description

Student: *Marlon E. Cobos*

GSoC Mentors: *Luis Osorio-Olvera, Vijay Barve, and Narayani Barve*

Motivation:

Ecological niche modeling represents a set of methodological tools that allow researchers to characterize and analyze species ecological niches. Currently, these methods are used widely and their applications include disease risk mapping, climate change risk predictions, conservation biology, among others. Physiological data suggests that Grinnellian ecological niches are convex in nature and they may probably have an ellipsoidal form when multiple dimensions are considered. However, among the diverse available software to characterize ecological niches, algorithms to model these niches as ellipsoids in the environmental space are scarce. Given the need for more solutions, the **ellipsenm** package aimed for developing specialized tools to perform multiple analyses related to ecological niches of species.

### Status of the project

At the moment we have completed the first of three phases. We have made few modifications to the list of original products that have helped us to improve the package functionality. Next steps are to complete scheduled activities to obtain all products.

All commits made can be seen at the <a href="https://github.com/marlonecobos/ellipsenm/commits/master" target="_blank">complete list of commits</a>.

Following you can find a brief description of this R package, as well as some examples of its use (still in development).

<br>

## Package description

The **ellipsenm** R package implements multiple tools to help in using ellipsoid envelopes to model ecological niches of species. Handly options for calibrating and selecting models, producing models with replicates and projections, and assessing niche overlap are included as part of this package. Other functions implemented here are useful to perform a series of pre- and post-modeling analyses.

<br>

## Installing the package

**ellipsenm** is in a GitHub repository and can be installed and/or loaded using the following code (make sure to have Internet connection).

Note: Try the following code first. If you have any problem during the installation, restart your session, close other R sessions you may have open, and try again. If during the installation you are asked to update packages, please do it (select the option that says All). If any of the packages gives an error, please install it alone using install.packages(), then try re-installing **ellipsenm** again. Also, it may be a good idea to update R and RStudio (if you are using it).

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# Installing and loading packages
if(!require(devtools)){
    install.packages("devtools")
}
if(!require(ellipsenm)){
    devtools::install_github("marlonecobos/ellipsenm")
}
library(ellipsenm)
```

<br>

## Using the package

### Modeling ecological niches using ellipsoids

An ecological niche, from a Grinnellian perspective, is the set of environmental conditions that allow a species to maintain populations for long periods of time, without immigration events. Models created with the ellipsenm package are ellipsoid envelope models and assume that a species ecological niche is convex, has an only optimum, and the species response to each variable covaries with the response to other variables. Mahalanobis distances are used to represent how far is each combination of environmental conditions from the optimum (ellipsoid centroid). Suitability values result by default from a multivariate normal transformation of Mahalanobis distances. Therefore, maximum values of suitability will be close to the centroid and minimum values will be close to the border of the ellipsoid.

We encourage the users to check the function's help before using it. This is possible using the code below:

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
help(ellipsoid_model)
```

Now the examples. Three distinct ways to create ecological niche models using **ellipsenm** are presented below:

<br>

**Creating a simple ecological niche model** 

When models are created this way, the whole dataset is used for fitting the ellipsoids.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# reading data
occurrences <- read.csv(system.file("extdata", "occurrences.csv",
                                    package = "ellipsenm"))

# raster layers of environmental data
vars <- raster::stack(list.files(system.file("extdata", package = "ellipsenm"),
                                 pattern = "bio", full.names = TRUE))

# creating the model
ell_model <- ellipsoid_model(data = occurrences, species = "species",
                             longitude = "longitude", latitude = "latitude",
                             raster_layers = vars, method = "covmat", level = 99,
                             replicates = 1, prediction = "suitability",
                             return_numeric = TRUE, format = "GTiff",
                             overwrite = FALSE, output_directory = "ellipsenm_model")

class(ell_model)
# check your directory, folder "ellipsenm_model"
```

<br>

**Modeling an ecological niche with replicates** 

When models are replicated, sub-samples of the data are used to create each replicate. Mean, minimum, and maximum ellipsoid models are also calculated. See more details about sub-sampling in the function's help.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# reading data
occurrences <- read.csv(system.file("extdata", "occurrences.csv",
                                    package = "ellipsenm"))

# raster layers of environmental data
vars <- raster::stack(list.files(system.file("extdata", package = "ellipsenm"),
                                 pattern = "bio", full.names = TRUE))

# creating the model with replicates
ell_model1 <- ellipsoid_model(data = occurrences, species = "species",
                              longitude = "longitude", latitude = "latitude",
                              raster_layers = vars, method = "covmat", level = 99,
                              replicates = 5, prediction = "suitability",
                              return_numeric = TRUE, format = "GTiff",
                              overwrite = FALSE, output_directory = "ellipsenm_model1")

class(ell_model1)
# check your directory, folder "ellipsenm_model1"
```

<br>

**Example of an ecological niche model with projections**

Ellipsoid envelope models can also be projected to other scenarios. This is one example using a RasterStack of layers. Projections can be done to multiple scenarios in same modeling process by changing the type of *projection_variables*. Check the function's help for more details.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# reading data
occurrences <- read.csv(system.file("extdata", "occurrences.csv",
                                    package = "ellipsenm"))

# raster layers of environmental data
vars <- raster::stack(list.files(system.file("extdata", package = "ellipsenm"),
                                 pattern = "bio", full.names = TRUE))

# creating the model with projections
pr_vars <- raster::stack(system.file("extdata", "proj_variables.tif",
                                     package = "ellipsenm"))
names(pr_vars) <- names(vars)

ell_model2 <- ellipsoid_model(data = occurrences, species = "species",
                              longitude = "longitude", latitude = "latitude",
                              raster_layers = vars, method = "covmat", level = 99,
                              replicates = 3, replicate_type = "bootstrap",
                              bootstrap_percentage = 75, projection_variables = pr_vars,
                              prediction = "suitability", return_numeric = TRUE,
                              format = "GTiff", overwrite = FALSE,
                              output_directory = "ellipsenm_model2")

class(ell_model2)
# check your directory, folder "ellipsenm_model2"
```

Other examples will be added soon.
